A forest is more than its trees: haplotypes and inferred ARGs

### 2024-09-03 paper/writeup.tex
notation change

% NOTE: these are in alphabetical order for now; we can revisit that ordering
%\author[$\dagger$]{Ava Bamforth}
\author[$\dagger$]{Halley Fritze}
\author[$\ddagger$]{Jerome Kelleher}
\author[$\dagger$]{Nathaniel Pope}
\author[$\ast$,$\dagger$,1]{Peter Ralph}

@@

% * description (HALLEY)
% * proof of something:
%     - arrives at a local minimum?
%     - guess at typical reduction? (reduces edges by 1/3?)
    We define a \textit{tree sequence}, denoted $\T$, 
    as a tuple $\left(T_1,...,T_{|\T|}\right)$ with a sequence of points
	$0 = a_0 \leq a_1 \leq \cdots < a_{|\T|} = L$ where
     each $T_k$ is a tree describing genealogical relationships between samples
     over a distinct portion of the genome $[a_{k-1}, a_k)$.
    Given any tree sequence, our goal is to
    identify areas of implied inheritance of haplotypes.
    We do so by identifying paths in pairs of consecutive tree sequences
    and merge them, given the correct condition.
    For example, if neighboring trees contain edges with 
    identical ancestor (parent) and descendant (child) nodes, 
    we assume that there is implied coalescence of haplotypes between nodes, 
    and we will extend the path's edges to the neighboring tree
     \ref{fig:extending_diagram}.
     More generally, for a pair of trees $T_k, T_{k+1}\in\mathbb{T}$,
     to transition from tree $T_k$ to tree $T_{k+1}$ we identify the 
     subforest of edges in $T_k$ that are removed to construct $T_{k+1}$
     and the subforest of edges in $T_{k+1}$ which were added to construct 
     $T_{k+1}$ from $T_k$. We call these subforest the out-forest and in-forest, respectively,
     and denoted $EO$ and $EI$. \comment{Need better notation. Currently cant think of anything better.}
     For each edge $e \in EO$ with child $c$,
     we construct a path $p_c^{EO}$ comprised of edges in the $EO$ subforest.
     If there exists an edge $e'\in EI$ also with child $c$,
     we construct path $p_c^{EI}$ from edges in the $EI$ subforest.
     If there is a common node in both paths excluding $c$, ie. $(p_c^{EO}\cap p_c^{EI})\backslash \{c\}\neq\emptyset$, then there is a mergable path between both subforests. The merged path $\bar{p}_c = \hat{p}_c^{EI} \cup \hat{p}^{EO}$ is the union of both paths truncated ($\hat{p}$) at the common intersection node. We then insert $\bar{p}$ into $T_2$, postponing or deleting edges in its original inforest $EI$, see Figure \ref{complicated extending diagram}. We perform this action across all consecutive pairs of tree sequences, going both forward and backward along the genome, until there are no more edges extended in this way.
%    We do so by identifying short paths (2 edges) in a tree.
%    \comment{Change this if we append new Extend Edges Method to account for longer paths $p\to x \to y\to c$}
%    If neighboring trees contain edges with 
%    identical ancestor (parent) and descendant (child) nodes, 
%    we assume that there is implied coalescence of haplotypes between nodes, 
%    and we will extend the path's edges to the neighboring tree \ref{fig:extending_diagram}.
%  	In more concrete terms, for a tree $T_k$ in the tree sequence $\T$ 
%    suppose there exists a path which contains a node $3$,
%    with parent and child nodes $4$ and $0$, respectively.
%    If there is an edge between the nodes $4$ and $0$ in trees 
%    $T_{k+1}$ (or $T_{k-1}$),
%    then we wish to extend the edges $4\to 3$ and $3\to 0$ 
%    into tree $T_{k+1}$ and then remove the edge $4 \to 0$ from $T_{k+1}$. 
%    This action reduces the length of $4 \to 0$ on the genome,
%    and in some cases, completely removes the edge. 
%    We now perform this action on all such ancestral paths 
%    over the entire tree sequence.
%    With algorithm \ref{alg:edge} we only extend existing edges
%    \comment{Note possible change pending algorithm change}
%    over a larger interval on the genome,
%    and remove unnecessary edges in the process.

\begin{figure}[!ht]
\begin{center}
	\includegraphics[width=5in]{edge_extend_method.pdf}
\end{center}
\caption{\comment{Change figure to say Extend halplotypes}A visualization of the \textit{extend haplotypes} method. Given two neighboring trees $T_1$ and $T_2$, $EO$ in the tree $T_1$ contains the edge $3\to 0$. Starting at $0$, there is a path of edges $0 \to 3 \to 4$. The tree $T_2$ has edge $4 \to 0$ in $EI$, and so there exists path $0 to 4$. These two paths share common end points, $0$ and $4$, and so we will extend path $0 \to 3 \to 4$ onto tree $T_2$, hence removing the edge $4\to 0$.
    \label{fig:extending_diagram}
}
\end{figure}

@@

	\caption{Extends haplotypes across a tree sequence $TS$.}\label{alg:hap}
	\SetKwFunction{ex}{Extend Paths}
	\SetKwFunction{fp}{Find Paths}
	\DontPrintSemicolon
	\LinesNotNumbered
	%\SetAlgoNoEnd
	\input{A tree sequence $\T$}
	\output{New tree sequence,$\T'$}
	\BlankLine
	\Fn{\fp (tree sequence $\T$, forwards = \tn{True})}{
	{
		\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
		{\BlankLine
			$\T = \left(T_{|\T|},..., T_1\right)$\;
		}
	}
		\For{pairs $(T_k, T_{k+1})\in \T\times\T$}{
		Compute subforests $EO\subseteq T_k$ and $EI\subseteq\in T_{k+1}$.\;
		\For{edges $e$ in $EO$}{
			Let $c$ be child node.\;
			$p_c^{EO} = [c]$\;
			$p_c^{EI} = [c]$\;
			$\bar{p}_c = []$ Path to be inserted\;
			\While{$\exists f\in EO$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))=0$}{
				$p = \tn{parent}(f)$\;
				Append $p$ to $p_c^{EO}$\;
				$c = p$\;
			}
			\While{$\exists f\in EI$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))<3$}{
				$p =\tn{parent}(f)$\;
				Append $p$ to $p_c^{EI}$\;
				$c = p$\;
			}
			\If{($p_c^{EO}\cap p_c^{EI})\backslash\{c\}\neq\emptyset$}{
				Merge paths up to intersection point\;
				$\bar{p}_c = \hat{p}_c^{EI} \cup \hat{p}^{EI}_c$\;
				and sort by node time.\;
			}
			Insert $\bar{p}_c$ into $T_{k+1}$\;		
		}
	}\;
	Update $\T$.\;
	\KwRet{$\T$}\;
}
\BlankLine						
\Fn{\ex ($\T$, iter=100)}{
	\For{$*$ \KwTo \Range(iter)}{
		$\T' \gets \T$.\fp(fowards=\tn{True})\;
		$\T' \gets \T$.\fp(fowards=\tn{False})\;
		\If{Number of edges in $\T$ = Number of edges in $\T'$}{
			\Break\BlankLine}
	}
	\KwRet{$\T'$}\;
}
\end{algorithm2e}

>>>>

% NOTE: these are in alphabetical order for now; we can revisit that ordering
\author[$\dagger$]{Ava Bamforth}
\author[$\dagger$]{Halley Fritze}
\author[$\ddagger$]{Jerome Kelleher}
\author[$\dagger$]{Nathaniel Pope}
\author[$\ast$,$\dagger$,1]{Peter Ralph}

@@

% * description (HALLEY)
% * proof of something:
%     - arrives at a local minimum?
%     - guess at typical reduction? (reduces edges by 1/3?)
    We define a \textit{tree sequence}, denoted $\T$, 
    as a tuple $\left(T_1,...,T_{|\T|}\right)$ with a sequence of points
	$0 = a_0 \leq a_1 \leq \cdots < a_{|\T|} = L$ where
     each $T_k$ is a tree describing genealogical relationships between samples
     over a distinct portion of the genome $[a_{k-1}, a_k)$.
    Given any tree sequence, our goal is to
    identify areas of implied inheritance of haplotypes.
    We do so by identifying paths in pairs of consecutive tree sequences
    and merge them, given the correct condition.
    For example, if neighboring trees contain edges with 
    identical ancestor (parent) and descendant (child) nodes, 
    we assume that there is implied coalescence of haplotypes between nodes, 
    and we will extend the path's edges to the neighboring tree
     \ref{fig:extending_diagram}.
     More generally, for a pair of trees $T_k, T_{k+1}\in\mathbb{T}$,
     to transition from tree $T_k$ to tree $T_{k+1}$ we identify the 
     subforest of edges in $T_k$ that are removed to construct $T_{k+1}$
     and the subforest of edges in $T_{k+1}$ which were added to construct 
     $T_{k+1}$ from $T_k$. We call these subforest the out-forest and in-forest, respectively,
     and denoted $F_O$ and $F_O$. \comment{Need better notation. Currently cant think of anything better.}
     For each edge $e \in F_O$ with child $c$,
     we construct a path $p_c^{F_O}$ comprised of edges in the $F_O$ subforest.
     If there exists an edge $e'\in F_I$ also with child $c$,
     we construct path $p_c^{F_I}$ from edges in the $F_I$ subforest.
     If there is a common node in both paths excluding $c$, ie. $(p_c^{F_O}\cap p_c^{F_I})\backslash \{c\}\neq\emptyset$, then there is a mergable path between both subforests. The merged path $\bar{p}_c = \hat{p}_c^{F_I} \cup \hat{p}^{F_O}$ is the union of both paths truncated ($\hat{p}$) at the common intersection node. We then insert $\bar{p}$ into $T_2$, postponing or deleting edges in its original inforest $F_I$, see Figure \ref{complicated extending diagram}. We perform this action across all consecutive pairs of tree sequences, going both forward and backward along the genome, until there are no more edges extended in this way.
%    We do so by identifying short paths (2 edges) in a tree.
%    \comment{Change this if we append new Extend Edges Method to account for longer paths $p\to x \to y\to c$}
%    If neighboring trees contain edges with 
%    identical ancestor (parent) and descendant (child) nodes, 
%    we assume that there is implied coalescence of haplotypes between nodes, 
%    and we will extend the path's edges to the neighboring tree \ref{fig:extending_diagram}.
%  	In more concrete terms, for a tree $T_k$ in the tree sequence $\T$ 
%    suppose there exists a path which contains a node $3$,
%    with parent and child nodes $4$ and $0$, respectively.
%    If there is an edge between the nodes $4$ and $0$ in trees 
%    $T_{k+1}$ (or $T_{k-1}$),
%    then we wish to extend the edges $4\to 3$ and $3\to 0$ 
%    into tree $T_{k+1}$ and then remove the edge $4 \to 0$ from $T_{k+1}$. 
%    This action reduces the length of $4 \to 0$ on the genome,
%    and in some cases, completely removes the edge. 
%    We now perform this action on all such ancestral paths 
%    over the entire tree sequence.
%    With algorithm \ref{alg:edge} we only extend existing edges
%    \comment{Note possible change pending algorithm change}
%    over a larger interval on the genome,
%    and remove unnecessary edges in the process.

\begin{figure}[!ht]
\begin{center}
	\includegraphics[width=5in]{edge_extend_method.pdf}
\end{center}
\caption{\comment{Change figure to say Extend halplotypes}A visualization of the \textit{extend haplotypes} method. Given two neighboring trees $T_1$ and $T_2$, $F_O$ in the tree $T_1$ contains the edge $3\to 0$. Starting at $0$, there is a path of edges $0 \to 3 \to 4$. The tree $T_2$ has edge $4 \to 0$ in $F_I$, and so there exists path $0 to 4$. These two paths share common end points, $0$ and $4$, and so we will extend path $0 \to 3 \to 4$ onto tree $T_2$, hence removing the edge $4\to 0$.
    \label{fig:extending_diagram}
}
\end{figure}

@@

	\caption{Extends haplotypes across a tree sequence $TS$.}\label{alg:hap}
	\SetKwFunction{ex}{Extend Paths}
	\SetKwFunction{fp}{Find Paths}
	\DontPrintSemicolon
	\LinesNotNumbered
	%\SetAlgoNoEnd
	\input{A tree sequence $\T$}
	\output{New tree sequence,$\T'$}
	\BlankLine
	\Fn{\fp (tree sequence $\T$, forwards = \tn{True})}{
	{
		\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
		{\BlankLine
			$\T = \left(T_{|\T|},..., T_1\right)$\;
		}
	}
		\For{pairs $(T_k, T_{k+1})\in \T\times\T$}{
		Compute subforests $F_O\subseteq T_k$ and $F_I\subseteq\in T_{k+1}$.\;
		\For{edges $e$ in $F_O$}{
			Let $c$ be child node.\;
			$p_c^{F_O} = [c]$\;
			$p_c^{F_I} = [c]$\;
			$\bar{p}_c = []$ Path to be inserted\;
			\While{$\exists f\in F_O$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))=0$}{
				$p = \tn{parent}(f)$\;
				Append $p$ to $p_c^{F_O}$\;
				$c = p$\;
			}
			\While{$\exists f\in F_I$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))<3$}{
				$p =\tn{parent}(f)$\;
				Append $p$ to $p_c^{F_I}$\;
				$c = p$\;
			}
			\If{($p_c^{F_O}\cap p_c^{F_I})\backslash\{c\}\neq\emptyset$}{
				Merge paths up to intersection point\;
				$\bar{p}_c = \hat{p}_c^{F_I} \cup \hat{p}^{F_I}_c$\;
				and sort by node time.\;
			}
			Insert $\bar{p}_c$ into $T_{k+1}$\;		
		}
	}\;
	Update $\T$.\;
	\KwRet{$\T$}\;
}
\BlankLine						
\Fn{\ex ($\T$, iter=100)}{
	\For{$*$ \KwTo \Range(iter)}{
		$\T' \gets \T$.\fp(fowards=\tn{True})\;
		$\T' \gets \T$.\fp(fowards=\tn{False})\;
		\If{Number of edges in $\T$ = Number of edges in $\T'$}{
			\Break\BlankLine}
	}
	\KwRet{$\T'$}\;
}
\end{algorithm2e}
